Generalizing the Model View Controller Paradigm

### 2006-08-10 inst/paper/paper2v2.tex
Made changes to paper 2.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19115 bc3139a8-67e5-0310-9ffc-ced21a209358

One particular model for creating interactive applications involves three
components: some user action or input, which is referred to as an event;
a response to that action, which is executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which is referred to as the signal handler.  The events can
range from the user moving the cursor over a plot, to pressing a button on a
graphical user interface (GUI), to pressing certain keys.  Events are
interpreted based on which window has focus.  For example, in the XWindows
system, if a user types C on the keyboard and the window that has focus is a
text editor, then the letter C is printed on the text editor's screen.
Any other open windows that do not have focus do not respond to this event.
The callback function is a unique function because its signature is determined
by the event that the callback function is linked to.  The parameters passed
to the callback function may include information about the event, such as the
on screen location of the event, and information about the object that had
focus when the event occurred, such as a button on the GUI.  The signal
handler is the piece that must look to see if an event has occurred and then
when the event has occurred, the signal handler calls the appropriate callback
function.

@@

In the design presented in this paper, the components consist of \Robject{MVC}
objects and communication within and between \Robject{MVC} objects is
performed by message objects.  This solution uses the power and simplicity of
the MVC design and expands it to the multiple data set problem.  Now we can
build on the original MVC design, which has been well tested, instead of
starting a new design from scratch.  The strength of this multiple linked MVC
design is that it uses MVC components, which can be individually created and
then linked.  Thus, these components are reusable.  These design decisions are
discussed further in Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.

@@

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views, and all views
must respond to a certain subset of events so we have implemented the view
classes by first creating a single, virtual class which the specialized
views extend.  As with our model classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

@@

Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects can be related to each other 
through a parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.
In this picture, modelA is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
modelB and modelC. Then from `modelB', three
\Robject{MVC} objects were created, modelD, modelE, and
modelF.  As an example, modelA could be gene 
expression data, modelB could be gene expression data that is a subset of
modelA (based on filtering), modelC could be derived by performing
multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
(GO) graph representing the molecular functions of the genes in modelB,
modelE could be the KEGG pathways that the genes in modelB are involved
in, and modelF could be the chromosome locations of the genes in modelB.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
    \caption{ Message passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatter plot (view 1).  Then an update data message is sent to
      the model in the second step, which alters the model.  Next in the third
      step all views of this model are sent an update view message, which
      causes the views to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

@@

Thinking of experimental data that are linked to meta-data as an example, there
are two \Robject{MVC} objects, one for the experimental data and one for the
meta-data.  In the example of experimental microarray data that are linked to
the Gene Ontology molecular function graph that was first mentioned in Section
\ref{Ssec:Limit}, these two models are linked in the following manner: the
Affymetrix identifiers in the experimental microarray data are linked to Locus
Link identifiers, which are in turn linked to the Gene Ontology molecular
function terms.  Thus, to pass information between these two models a table
that links Affymetrix identifiers to Gene Ontology molecular function terms
must be stored or there must be a function to call that converts from
Affymetrix identifiers to Gene Ontology molecular function terms and vice
versa.  Once it is possible to link these models, then information must be
passed between them through messages to keep the views synchronized.

@@

When a child \Robject{MVC} is created from an existing \Robject{MVC}, the
child model's \Rslot{linkData} slot is filled with the two functions,
\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in
the child \Robject{MVC} is updated, the \Rfunction{toParent} function in the
\Rslot{linkData} slot is used to convert the child model's data
change into useful information for the parent model.

Now suppose that the parent model changed instead.  This information
needs to be passed to the child model.  However, a parent can have
multiple child \Robject{MVC} objects so the parent sends a message to each
child \Robject{MVC} to notify the child of the change and then each child
\Robject{MVC} handles the message by using its \Rfunction{fromParent}
function.  The \Rfunction{fromParent} function converts the parent
model's data change into useful information for this
particular child model.  Thus, the message sent to all children is the
same, but each child converts the message, using its \Rfunction{fromParent}
function, to make it specific for that child.

>>>>

One particular model for creating interactive applications involves three
components: some user action or input, which is referred to as an event;
a response to that action, which is executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which is referred to as the signal handler.  The events can
range from the user moving the cursor over a plot, to pressing a button on a
graphical user interface (GUI), to pressing certain keys.  Events are
interpreted based on which window has focus.  For example, in the X Windows
system, if a user types C on the keyboard and the window that has focus is a
text editor, then the letter C is printed on the text editor's screen.
Any other open windows that do not have focus do not respond to this event.
The callback function is a unique function because its signature is determined
by the event that the callback function is linked to.  The parameters passed
to the callback function may include information about the event, such as the
on screen location of the event, and information about the object that had
focus when the event occurred, such as a button on the GUI.  The signal
handler is the piece that must look to see if an event has occurred and then
when the event has occurred, the signal handler calls the appropriate callback
function.

@@

In the design presented in this paper, the components consist of \Robject{MVC}
objects, and communication within and between \Robject{MVC} objects is
performed by message objects.  This solution uses the power and simplicity of
the MVC design and expands it to the multiple data set problem.  Now we can
build on the original MVC design, which has been well tested, instead of
starting a new design from scratch.  The strength of this multiple linked MVC
design is that it uses MVC components, which can be individually created and
then linked.  Thus, these components are reusable.  These design decisions are
discussed further in Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.

@@

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views, and all views
must respond to a certain subset of events so we have implemented the view
classes by first creating a single, virtual class, which the specialized
views extend.  As with our model classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

@@

Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects can be related to each other 
through a parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.
In this picture, modelA is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
modelB and modelC. Then from modelB, three
\Robject{MVC} objects were created, modelD, modelE, and
modelF.  As an example, modelA could be gene 
expression data, modelB could be gene expression data that is a subset of
modelA (based on filtering), modelC could be derived by performing
multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
(GO) graph representing the molecular functions of the genes in modelB,
modelE could be the KEGG pathways that the genes in modelB are involved
in, and modelF could be the chromosome locations of the genes in modelB.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMPwithin2.jpg}}
    \caption{ Message passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatter plot (view 1).  Then an update data message is sent to
      the model in the second step, which alters the model.  Next in the third
      step all views of this model are sent an update view message, which
      causes the views to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

@@

Thinking of experimental data that are linked to meta-data as an example, there
are two \Robject{MVC} objects, one for the experimental data and one for the
meta-data.  In the example of experimental microarray data that are linked to
the Gene Ontology molecular function graph that was first mentioned in Section
\ref{Ssec:Limit}, these two models are linked in the following manner: the
Affymetrix identifiers in the experimental microarray data are linked to Entrez
Gene identifiers, which are in turn linked to the Gene Ontology molecular
function terms.  Thus, to pass information between these two models a table
that links Affymetrix identifiers to Gene Ontology molecular function terms
must be stored or there must be a function to call that converts from
Affymetrix identifiers to Gene Ontology molecular function terms and vice
versa.  Once it is possible to link these models, then information must be
passed between them through messages to keep the views synchronized.

@@

When a child \Robject{MVC} is created from an existing \Robject{MVC}, the
child model's \Rslot{linkData} slot is filled with the two functions,
\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in
the child \Robject{MVC} is updated, the \Rfunction{toParent} function in the
\Rslot{linkData} slot is used to convert the child model's data
change into useful information for the parent model.  
Now suppose that the parent model changed instead.  This information
needs to be passed to the child model.  However, a parent can have
multiple child \Robject{MVC} objects so the parent sends a message to each
child \Robject{MVC} to notify the child of the change and then each child
\Robject{MVC} handles the message by using its \Rfunction{fromParent}
function.  The \Rfunction{fromParent} function converts the parent
model's data change into useful information for this
particular child model.  Thus, the message sent to all children is the
same, but each child converts the message, using its \Rfunction{fromParent}
function, to make it specific for that child.
